---
title: "Activities"
description: |
  Go ahead, don´t be shy!
toc: TRUE
toc_float: TRUE
site: distill::distill_website
---

# Activities

## Activity 1

Open up R Studio and in the console, type the following code:

```{r, echo = TRUE, eval = FALSE}
?rnorm
```

The help documentation for `rnorm()` should appear in the bottom right help panel. In the usage section, we see that `rnorm()` takes the following form:

```{r, echo = TRUE, eval = FALSE}
rnorm(n, mean = 0, sd = 1)
```

In the arguments section, there are explanations for each of the `arguments`. `n` is the number of observations we want to create, `mean` is the mean of the data points we will create and `sd` is the standard deviation of the set. In the details section it notes that if no values are entered for `mean` and `sd` it will use a default of 0 and 1 for these values. Because there is no default value for `n` it must be specified otherwise the code won't run.

Let's try an example and just change the required argument `n` to ask R to produce 5 random numbers.

## Activity 2

Copy and paste the following code into the console.

```{r, echo = TRUE, eval = FALSE}
set.seed(12042016)
rnorm(n = 5, mean = 10, sd = 2)
```

This time R has still produced 5 random numbers, but now this set of numbers has a mean of 10 and an sd of 2 as specified. 

Try to find out which `functions` allow you to calculate the mean or the standard deviation of a column (or "variable").

Always remember to use the help documentation to help you understand what arguments a function requires.

## Activity 3
Open up RStudio and create a new project file.

After creating your new project, create a new R script and copy and paste the following code into the script. Save the script.

```{r, echo = TRUE, eval = FALSE}
set.seed(12042016)
norm_distb <- rnorm(n = 600, mean = 10, sd = 2)

hist(norm_distb)
```

## Activity 4
Create a new RMarkdown file in your project folder. Change all the setting in the YAML-header (i.e., your name.)

* We always the date to be up-to-date: Try to replace the set date by a function we have used before (hint: your computer like to tell you)

* Add the following options to the YAML-header

```{r, echo = TRUE, eval = FALSE}
output: 
  bookdown::html_document2:
    toc: TRUE
    toc_float: TRUE
```

* Add a new headline to the RMarkdown file (indicated by the `#`) and copy and paste the following code beneath (set `eval =` to `TRUE`)

```{r, echo = TRUE, eval = FALSE}
set.seed(12042016)
norm_distb <- rnorm(n = 600, mean = 10, sd = 2)

hist(norm_distb)
```

* Render! (knit)

## Activity 5
Re-open your RMarkdown file and add a new code `chunk`. Try to import one of the datasets you can find via the `datasets` tab.

Create a histogram (as in Activity 4) for one of the columns.

## Activity 6
Is part of the introductory post on grammar of ggplot2.


## Activity 7
Is part of the introductory post on grammar of ggplot2.

## Activity 8


